---
title: "Reproducible Research: Peer Assessment 1"
author: "Kal Elsidawi"
date: "`r Sys.Date()`"
output: 
        html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Sourcing the assignment instructions and the related raw/processed dataset 
After forking the assignment from GitHub, find below the list of files:

Note: Make sure that your working directory is set appropriately to the 
root of the repository on your local disk

```{r}
list.files(recursive = TRUE)
rm(list = ls())
```
Note that the instructions are found in the **instructions.pdf** file under the **doc** directory
The dataset is in the **activity.zip** file

## Loading and preprocessing the data
The dataset is stored in a tibble called *activity* 

```{r}
library(data.table)
library(tibble)
library(tidyverse)
library(lattice)
unzip("activity.zip", list = TRUE)
activity <- as_tibble(read.table(file = unzip("activity.zip", "activity.csv"), header = TRUE,sep = ",",
                                 na.strings = NA, colClasses = c("numeric","Date","numeric") ))

```

## What is mean total number of steps taken per day?

First of all, create a new data frame **dailySteps** that has TotalSteps taken
per day then create a histogram

```{r}
dailySteps <- activity %>% group_by(date) %>% summarise(TotalSteps = sum(steps, na.rm = TRUE))
breaks <- seq(from = min(dailySteps$TotalSteps),
              to = max(dailySteps$TotalSteps),
              length.out = 20
)
hist(dailySteps$TotalSteps,main = 'Histogram of TotalSteps per day',
     breaks = breaks,col = 'lightblue',
     xlab = 'Total Steps per day'
)

```

The **mean** and **median** of the total number of steps taken each day:

```{r}
summarytbl <- summary(dailySteps$TotalSteps)
TS_mean <- summarytbl['Mean']
TS_median <- summarytbl['Median']
summarytbl
```

## What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis)

```{r}
dailyIntervalSteps <- activity %>% group_by(interval) %>% summarise(AvgDailyIntervalSteps = mean(steps, na.rm = TRUE))
with(dailyIntervalSteps,plot(x = interval, y = AvgDailyIntervalSteps,
                             type = 'l', col ='blue',
                             main = "Average # of Steps during 24:00 hours"
)
)
```
To find the 5-min interval in which the maximum number of steps take place
across all days, we search for the interval whose average steps over the days is
the maximum across all other intervals
This comes out to be the 5 minutes between 8:35 a.m. and 8:40 a.m. which is
interval 104 out of the 288 intervals per day ( 24hrs x 12 5min-interval/hr = 288)

```{r}
dailyIntervalSteps[which.max(dailyIntervalSteps$AvgDailyIntervalSteps),]

```



## Imputing missing values
The total number of missing values in the field *steps* in the dataset **activity**
is calculated by first creating a logical vector that computes a TRUE when steps
value is NA and FALSE otherwise. Then uses the table function that reports the
sum of all FALSE( not NA) and TRUE (NA) entries as shown below

```{r}

stepsV <- activity$steps
missing <- is.na(stepsV)
table(missing)
# indecies of missing values
indexOfNA <- which(missing == TRUE)
```

Therefore, there are 2304 missing values out of a total of 17,568 records
that is ```{r} paste(round((2304/17568)*100,2), '%')```

The missing value in a specific 5-min interval will be replaced by the mean
of the 5-min interval calculated across all days

To do that, we will use the logical vector previously calculated **missing** 
in order to identify which row / observation where the steps needs to be filled by the 
corresponding similar 5-min interval slot from table **dailyIntervalSteps**
The indices are stored in **indexOFNA**

---

The trick is to calculate the modulo of the interval number by 288 (which is the 
remainder of the division by 288). Each 288 intervals represent a full day and 
the remainder would then provide the interval number of that observation within 
that day. So rows 289 to 576 should correspond to 1 to 288. 
So if you take rows from 1 to 288, you need to get corresponding 1 to 288.
Using modulo 288 will achieve that objective except that it wouldn't work for 288
as you will get a remainder of 0. Consequently, 288 and any multiple of it will
be replaced by 288 when extracting the average value of steps for that interval;
which is the last 5 minute interval in a day.
Moreover, we will make a copy of the activity table and fill the NAs there in
order not to modify the raw dataset. We will call the copy : **activity_imu** 
i.e. activity with imputed missing values

```{r}
activity_imv <- activity
# find all the observation row numbers where the steps are NA
indexOfNA <- which(missing == TRUE)
# cycle through a loop of all the records with steps -> NA and replace their 
# value with the corresponding average for the same time slot from dailyIntervalSteps
```

Missing values before imputing:
```{r}
table(is.na(activity_imv$steps))
```

Replacing the NA values:

```{r}
i <- 0L
for (r in indexOfNA) {
        rSlot <- if(as.integer(r %% 288) != 0){
                as.integer(r %% 288) 
        } else {
                288
        }
        i <- i +1L
        activity_imv[as.integer(r),'steps'] <- dailyIntervalSteps[rSlot,'AvgDailyIntervalSteps']
        # uncomment below code if you would like to see the first 300 rows how their NA were filled
        # if (i < 300) { print(paste('r=',r,' rSlot =',rSlot,'activity steps filled by =',activity_imv[as.integer(r),'steps']) )
        # } else if(i == 301){
        #                 print(activity_imv[1:301,], n = 301)
        #         }
}
```
NA steps of the 2304 records have been filled. If you want to verify, please uncomment the print statement shown below:

```{r}
# print(activity_imv[indexOfNA,],n = 2304)
```


Missing values after imputing are not present as shown below:

```{r}
table(is.na(activity_imv$steps))
```

## What is mean total number of steps taken per day - after the missing data has been replaced?

Similarly as above, create a new data frame **dailySteps_imv** that has TotalSteps taken
per day then create a histogram
Producing the 

```{r}
dailySteps_imv <- activity_imv %>% group_by(date) %>% summarise(TotalSteps = sum(steps))
breaks <- seq(from = min(dailySteps_imv$TotalSteps),
              to = max(dailySteps_imv$TotalSteps),
              length.out = 20
)
hist(dailySteps_imv$TotalSteps,main = 'Histogram of TotalSteps per day',
     breaks = breaks,col = 'lightblue',
     xlab = 'Total Steps per day'
)

```

### the new mean and median of the adjusted data set after replacing the NAs:

```{r}
summarytbl_imv <- summary(dailySteps_imv$TotalSteps)
TS_mean_imv <- summarytbl_imv['Mean']
TS_median_imv <- summarytbl_imv['Median']
summarytbl_imv
```

The impact of the existance of NA value has skewed the Mean and Median as follows:

**Mean**

`r paste('The mean has ',if( (TS_mean - TS_mean_imv) <= 0){'decreased'} else {'increased'},' from  ',round(TS_mean_imv,2),' To ',round(TS_mean, 2), '\n i.e. ',round((abs(TS_mean - TS_mean_imv)/TS_mean_imv)*100, 2),' %\n')`

**Median**

`r paste('The median has ',if( (TS_median - TS_median_imv) <= 0){'decreased'} else {'increased'},'from ',round(TS_median_imv,2),' To ',round(TS_median,2),'\n i.e. ',round((abs(TS_median - TS_median_imv)/TS_median_imv)*100, 2),' %\n')`


## Are there differences in activity patterns between weekdays and weekends?

First build two vector lists to identify workdays and weekend as follows
Then mutate the dataset, adding another factor vector called *work* with labels
workday and weekend. Use the weekdays function comparing whether in the workdays
list to create the factor

``` {r}

workdays <- c("Monday" ,"Tuesday","Wednesday", "Thursday", "Friday")
weekend  <- c("Saturday", "Sunday")

activity_imv <-  activity_imv %>% mutate(work = factor((weekdays(date) %in% workdays), labels = c('weekday','weekend')))

```
Now plot the average number of steps per 5-min intervals in a day in a panel plot
comparing between weekday and weekend results

```{r}
dailyIntervalStepsByWork <- activity_imv %>% select(steps, date, interval, work) %>%  group_by(interval, work) %>% summarise(AvgDailyIntervalSteps = mean(steps, na.rm = TRUE))
dailyIntervalStepsByWork %>%
        xyplot(x = AvgDailyIntervalSteps ~ interval| work, groups = work,
               type = 'l',
               ylab = "Number of Steps",
               grid = TRUE,
               layout = c(1,2)
               )
#               ylab = 'Number of Steps', 
#               main = 'Comparing activity during workdays vs weekend',
               

```
# Summary

---

During weekdays on average between 10 a.m. and 8:30 p.m. activity varies 
between 50 and 100 steps per 5-min interval; whereas durign that same period
on the weekends, the activity is much less and varies between 25 and 75 steps 
per 5-min interval

Interestingly and in contrast, in the period between 8 and 9 a.m. activity is
higher during the weekend (150 to 200 steps) whereas during weekdays varies 
between 75 and 150 steps.



