---
title: "Reproducible Research: Peer Assessment 1"
author: "Kal Elsidawi"
date: "`r Sys.Date()`"
output: 
        html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Sourcing the assignment instructions and the related raw/processed dataset 
After forking the assignment from GitHub, find below the list of files:

```{r}
list.files(recursive = TRUE)
```
Note that the instructions are found in the **instructions.pdf** file under the **doc** directory
The dataset is in the **activity.zip** file

## Loading and preprocessing the data
The dataset is stored in a tibble called *activity* 
```{r}
library(data.table)
library(tibble)
library(tidyverse)
unzip("activity.zip", list = TRUE)
activity <- as_tibble(read.table(file = unzip("activity.zip", "activity.csv"), header = TRUE,sep = ",",
                       na.strings = NA, colClasses = c("numeric","Date","numeric") ))

```

## What is mean total number of steps taken per day?

First of all, create a new data frame **dailySteps** that has TotalSteps taken
per day then create a histogram

```{r}
dailySteps <- activity %>% group_by(date) %>% summarise(TotalSteps = sum(steps, na.rm = TRUE))
hist(dailySteps$TotalSteps)

```

Histogram of the total number of steps taken each day
## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
